---
title: "Final project - Part 2"
author: "Raphael Carvalho"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this paper, we're going to analyze the ToothGrowth data in the R datasets package.

1. Load the ToothGrowth data and perform some basic exploratory data analyses
2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.
4. State the conclusions and the assumptions needed for the conclusions.


### Loading the packages 

```{r packages, echo = TRUE, warning = FALSE, message=FALSE}
  library(datasets)
  library(dplyr)
  library(ggplot2)
```

#### Getting a glimpse of the data:

```{r glimpse, echo = FALSE, results='hold'}
  glimpse(ToothGrowth)
```

```{r summary, echo = TRUE, results='hold'}
  print(paste("Rows with NA values:", sum(!complete.cases(ToothGrowth))))
  print(summary(ToothGrowth))
```

### Exploring the dataset

```{r dispersion_dose, echo = FALSE, results='hold'}

ToothGrowth$dose <- as.factor(ToothGrowth$dose)
d1 <- ggplot(ToothGrowth, aes(x=dose, y=len,fill=dose)) +  geom_boxplot() +  ggtitle("Dispersion of tooth growth by dose") + xlab("Dose in mg") + ylab("Tooth length") + theme_minimal()
print(d1)
```

We can see that the longer theeth received higher doses. We can also notice that the dispersion of teeth length that were given a 1 mg dose is very high compared to the teeth subimitted to other doses. Lastly, we can see an outlier tooth at the group of 0.5 mg. 

```{r dispersion_delivery, echo = FALSE, results='hold'}

d2 <- ggplot(ToothGrowth, aes(x=supp, y=len,fill=supp)) +  geom_boxplot() +  ggtitle("Dispersion of tooth growth by delivery method") + xlab("Delivery method") + ylab("Tooth length") + theme_minimal()
d2
```
The tooth length is similar for both delivery methods. However, we can see that the dispersion for Orange Juice is much lower than the dispersion observed for Vitamin C.

### Testing some hyphotesis

#### Setting the variables:
```{r variables, echo = TRUE, results='hold'}

dose <- ToothGrowth$dose
supp <- ToothGrowth$supp
len <-  ToothGrowth$len

```

#### Does the delivery mode has an influence on the tooth growth?
```{r test_delivery, echo = TRUE, results='hold'}
t.test(len[supp == "VC"],len[supp == "OJ"], paired=FALSE)
```

**This test shows us that:**

1. We shouldn't reject the null hypothesis as the p-value is bigger than .05.
2. The confidence interval contains 0, so the test is not really significant.

#### Does the dose mode has an influence on the tooth growth?
```{r test_dose, echo = TRUE, results='hold'}
t.test(len[dose == 0.5],len[dose == 1], paired=FALSE)
```
**This test shows us that:**

1. We should reject the null hypothesis as the p-value is almost 0.
2. The confidence interval does not contain 0, so the test is significant.


### Conclusion

1. The dose of Vitamin C is proved to be a factor of teeth's growth for Guinea pigs;
2. The delivery mode does not appear to have an effect on teeth's growth.
